Packing limits and developmental integration are not sufficient to explain stomatal anatomical evolution in flowering plants

### 2022-09-11 ms/ms.tex
minor revisions

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% \documentclass[11pt]{article} % redundant
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\usepackage{multirow} % for merging rows in tables

@@

\author{}
\date{\vspace{-2.5em}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\textit{Word Count}:

@@

Functional and developmental constraints on phenotypic variation may cause traits to covary over millions of years and slow populations from reaching their adaptive optima. Alternatively, trait covariation may result from selective constraint if some trait combinations are generally maladaptive. Quantifying the relative contribution of functional, developmental, and selective constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. The stomatal anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the significance of functional and developmental constraints on phenotypyic evolution at broad phylogenetic scales. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially very different selective constraints. If stomatal traits on each surface evolve differently, this implies that independent evolution is possible and functional or developmental constraints alone likely do not explain trait covariance. Packing limits and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. We tested this by synthesizing data on stomatal density and length from amphistomatous leaves of 638 terrestrial flowering plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from 236 phylogenetically independent contrasts using a robust Bayesian model. Contrary to packing limit and developmental integration hypotheses, stomatal anatomy on each surface diverged partially independently. To better understand the global variation in ecologically important traits like stomata will likely require adaptive explanations for what constrains phenotypic (co)variation.

@@

The ability for traits to evolve independently of one another is a necessary prerequisite for adaptation to complex environments (\protect\hyperlink{ref-lewontin_adaptation_1978}{Lewontin 1978}). If traits can evolve independently and there is sufficient genetic variation, then selection should move populations toward their multivariate phenotypic optimum. Yet divergence in one trait often covaries with other traits and covariance can persist for millions of years (\protect\hyperlink{ref-schluter_adaptive_1996}{Schluter 1996}). Covariance is ``a rough local measure of the strength of constraint'' (\protect\hyperlink{ref-maynard_smith_developmental_1985}{Maynard Smith et al. 1985}) that can be broken down into functional, developmental, and selective constraints. Functional constraints are ``limitations imposed by time, energy, or the laws of physics'' (\protect\hyperlink{ref-arnold_constraints_1992}{Arnold 1992}). In other words, certain trait combinations are not physically or geometrically possible. A classic example is shell coiling among invertebrate lineages in which the morphospace of possible phenotypes is constrained by hard geometrical limits (\protect\hyperlink{ref-raup_geometric_1966}{Raup 1966}; \protect\hyperlink{ref-mcghee_theoretical_1999}{McGhee 1999}). Within the space of possible phenotypes, developmental constraints can ``bias\ldots the production of variant phenotypes or {[}place{]} a limitation on phenotypic variability'' (\protect\hyperlink{ref-maynard_smith_developmental_1985}{Maynard Smith et al. 1985}). For example, Fibonacci phyllotaxis may arise from a packing constraint of primordia on the developing apex (\protect\hyperlink{ref-mitchison_phyllotaxis_1977}{Mitchison 1977}; \protect\hyperlink{ref-reinhardt_law_2022}{Reinhardt and Gola 2022}; but see \protect\hyperlink{ref-niklas_role_1988}{Niklas 1988} for an adaptive explanation). Natural selection constrains phenotypic variation by preventing maladaptive forms from evolving. Selection causes trait covariation because `missing' trait combinations are maladaptive. Understanding phenotypic constraint is challenging, but a useful starting point is determining whether phenotypic covariation can be explained by functional or developmental constraints (\protect\hyperlink{ref-mcghee_theoretical_1999}{McGhee 1999}, \protect\hyperlink{ref-mcghee_geometry_2007}{2007}; \protect\hyperlink{ref-olson_plant_2019}{Olson 2019}). If phenotypic covariation is inconsistent with functional and developmental constraints, this provides a strong impetus to test for selective constraint.

In this study, we will address packing problems and developmental integration, specific forms of functional and developmental constraint relevant to our study system, stomatal anatomy. We introduce these concepts generally in this paragraph. Packing a number objects into a finite space is a common functional constraint on organisms. Regular geometries that appear in nature such as helices and hexagons (think DNA and honeycombs) are often optimal solutions to packing problems (\protect\hyperlink{ref-mackenzie_proving_1999}{Mackenzie 1999}; \protect\hyperlink{ref-maritan_optimal_2000}{Maritan et al. 2000}). Notice that functional constraint does not preclude selection, but the presence of a packing limit changes the range of possible phenotypes. Developmental integration is a form of developmental constraint on multivariate phenotypic evolution and we use them interchangeably in this study. Developmentally integrated traits have a ``disposition for covariation'' (\protect\hyperlink{ref-armbruster_integrated_2014}{Armbruster et al. 2014}), meaning that evolutionary divergence between lineages in one trait will be tightly associated with divergence in another trait. Allometry is a classic, albeit contested, example of developmental integration that may constrain phenotypic evolution (reviewed in \protect\hyperlink{ref-pelabon_evolution_2014}{Pélabon et al. 2014}). Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection (\protect\hyperlink{ref-lande_quantitative_1979}{Lande 1979}). However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection (\protect\hyperlink{ref-hansen_is_2003}{Hansen 2003}). For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators (Berg's rule, \protect\hyperlink{ref-berg_general_1959}{Berg 1959}, \protect\hyperlink{ref-berg_ecological_1960}{1960}; \protect\hyperlink{ref-conner_raissa_2014}{Conner and Lande 2014}; \protect\hyperlink{ref-armbruster_covariance_1999}{Armbruster et al. 1999}).

Biologists have studied phenotypic constraints for decades, but progress is challenging because many multivariate phenotypes are too complex or poorly understood to quantitatively distinguish functional, developmental, and selective constraints over macroevolutionary timescales. Stomatal anatomy on the leaves of flowering plants provide an exceptional opportunity because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of years of evolutionary history. For this purpose, a heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selective constraints may differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selective constraints most likely explain at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological significance. The next sections provide background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important. Readers not interested in this background may skip ahead to hypotheses and predictions.

@@

Stomata are microscopic pores formed by a pair of guard cells that regulate gas exchange (CO\(_2\) gain and water vapor loss) on the leaf or other photosynthetic surfaces of most land plants. Stomata originated once in the history of land plants around 500 Ma, diversified rapidly in density and size, and have been maintained in most lineages except some bryophytes and aquatic plants (recently reviewed in \protect\hyperlink{ref-clark_origin_2022}{Clark et al. 2022}). Stomata respond physiologically by opening and closing in response to light, humidity, temperature, circadian rhythm, and plant water status (\protect\hyperlink{ref-hetherington_role_2003}{Hetherington and Woodward 2003}; \protect\hyperlink{ref-lawson_guard_2020}{Lawson and Matthews 2020}). On this short time scale, the stomatal size, density, and distribution on a mature leaf do not change, so the maximum rate of is fixed. At somewhat longer timescales, the plant may respond to environmental cues such as light and CO\(_2\) by altering stomatal anatomy in new leaves (\protect\hyperlink{ref-casson_influence_2008}{Casson and Gray 2008}). Physiological responses (aperture change) and plastic responses (new leaves with changed anatomy) may be alternative strategies for plants to acclimate to environmental change (\protect\hyperlink{ref-haworth_co-ordination_2013}{Haworth, Elliott-Kingston, and McElwain 2013}). Finally, stomatal anatomy can evolve due to inherited changes in stomatal development. Plastic and genetic changes in stomatal anatomy are both ecologically important, but most studies do not use a common garden design that would tease apart their relative contribution.

We focus on anatomical variation in the density, size, and patterning of stomata on a leaf because these factors set the maximum stomatal conductance to CO\(_2\) diffusing into a leaf and the amount of water that transpires from it (\protect\hyperlink{ref-sack_hydrology_2003}{Sack et al. 2003}; \protect\hyperlink{ref-franks_effect_2001}{Franks and Farquhar 2001}; \protect\hyperlink{ref-galmes_leaf_2013}{Galmés et al. 2013}; \protect\hyperlink{ref-harrison_influence_2020}{Harrison et al. 2020}). Plants typically operate below their anatomical maximum by dynamically regulating stomatal aperture. Even though operational stomatal conductance determines the realized photosynthetic rate and water-use efficiency, anatomical parameters are useful in that they set the range of stomatal function (\protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}) and are correlated with actual stomatal function under natural conditions (\protect\hyperlink{ref-murray_consistent_2020}{Murray et al. 2020}). All else being equal, larger, more densely packed, but evenly spaced stomata increase gas exchange (\protect\hyperlink{ref-franks_maximum_2009}{Franks and Beerling 2009}; \protect\hyperlink{ref-dow_physiological_2014}{Dow, Berry, and Bergmann 2014}; \protect\hyperlink{ref-lehmann_effects_2015}{Lehmann and Or 2015}). Smaller stomata may also be able to respond more rapidly than larger stomata, proving the ability of leaves to track short duration environmental change (\protect\hyperlink{ref-drake_smaller_2013}{Drake, Froend, and Franks 2013}). Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in some species (\protect\hyperlink{ref-metcalfe_anatomy_1950}{Metcalfe and Chalk 1950}; \protect\hyperlink{ref-parkhurst_adaptive_1978}{Parkhurst 1978}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}). Amphistomatous leaves have a second parallel pathway from the substomatal cavities through the leaf internal airspace to sites of carboxylation in the mesophyll (\protect\hyperlink{ref-parkhurst_adaptive_1978}{Parkhurst 1978}; \protect\hyperlink{ref-gutschick_photosynthesis_1984}{Gutschick 1984}). Thus amphistomatous leaves have lower resistance to diffusion through the airspace which increases the photosynthetic rate (\protect\hyperlink{ref-parkhurst_intercellular_1990}{Parkhurst and Mott 1990}). If total stomatal and other conductances to CO\(_2\) supply could be held constant, then an amphistomatous leaf will have a greater conductance than am otherwise identical hypostomatous leaf. The magnitude of the advantage depends on upon the resistance to diffusion through the internal airspace, which is variable among species.

The adaptive significance and ecological distribution of leaves with different stomatal anatomies is complex and there is much yet to learn. Seed plants posses a wider range of stomatal anatomies than ferns and fern allies, which are restricted to having large stomata, at low density, only on the lower surface (\protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}). In general, trees and shrubs have greater stomatal density than herbs, but there is a of lot variation within growth forms depending on the ecological niche (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-kelly_plant_1995}{Kelly and Beerling 1995}). A commonly observed trend is that leaves from higher light environments tend to have greater stomatal density (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-gibson_structure-function_1996}{Gibson 1996}). This may explain why, perhaps counterintuitively, plants in dry environments tend to have more stomata. Drier habitats are more open, enabling plants with higher stomatal density can photosynthesize more when water is available, but close stomata during drought (\protect\hyperlink{ref-liu_variation_2018}{Liu et al. 2018}). Over recent human history, stomatal density has tended to decline within species as atmospheric CO\(_2\) concentrations have risen (\protect\hyperlink{ref-woodward_stomatal_1987}{Woodward 1987}; \protect\hyperlink{ref-royer_stomatal_2001}{Royer 2001}). It is unclear whether most of this change is plastic or genetic, but the overall direction is consistent with the hypothesis that plants decrease gas exchange as CO\(_2\) availability increases.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% \documentclass[11pt]{article} % redundant
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\usepackage{multirow} % for merging rows in tables

@@

\author{}
\date{\vspace{-2.5em}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\textit{Word Count}: 4,704

@@

Functional and developmental constraints on phenotypic variation may cause traits to covary over millions of years and slow populations from reaching their adaptive optima. Alternatively, trait covariation may result from selective constraint if some trait combinations are generally maladaptive. Quantifying the relative contribution of functional, developmental, and selective constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. The anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the importance of functional and developmental constraints on phenotypyic evolution. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially different selective constraints because of leaf asymmetry in light capture, gas exchange, and other features. Independent evolution of stomatal traits on each surface imply that functional and developmental constraints alone likely do not explain trait covariance. Packing limits on how many stomata can fit into a finite epidermis and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. The simple geometry of the planar leaf surface and knowledge of stomatal development make it possible to derive equations for phenotypic (co)variance caused by these constraints and compare them with data. We analyzed evolutionary covariance between stomatal density and length in amphistomatous leaves from 236 phylogenetically independent contrasts using a robust Bayesian model. Stomatal anatomy on each surface diverges partially independently, meaning that packing limits and developmental integration are not sufficient to explain phenotypic (co)variation. Hence, selective constraints, which require an adaptive explanation, likely contribute to (co)variation in ecologically important traits like stomata. We show how it is possible to evaluate the contribution of different constraints by deriving expected patterns of (co)variance and testing them using similar but separate tissues, organs, or sexes.\\

@@

The ability for traits to evolve independently of one another is a necessary prerequisite for populations to ascend a fitness peak in multivariate trait space (\protect\hyperlink{ref-lewontin_adaptation_1978}{Lewontin 1978}). If traits can evolve independently and there is sufficient genetic variation, then selection should move populations toward their multivariate phenotypic optimum. Yet divergence in one trait often covaries with other traits and covariance can persist for millions of years (\protect\hyperlink{ref-schluter_adaptive_1996}{Schluter 1996}). Covariance is ``a rough local measure of the strength of constraint'' (\protect\hyperlink{ref-maynard_smith_developmental_1985}{Maynard Smith et al. 1985}) that can be broken down into functional, developmental, and selective constraints. Functional constraints are ``limitations imposed by time, energy, or the laws of physics'' (\protect\hyperlink{ref-arnold_constraints_1992}{Arnold 1992}). In other words, certain trait combinations are not physically or geometrically possible. A classic example is shell coiling among invertebrate lineages in which the morphospace of possible phenotypes is constrained by hard geometrical limits (\protect\hyperlink{ref-raup_geometric_1966}{Raup 1966}; \protect\hyperlink{ref-mcghee_theoretical_1999}{McGhee 1999}). Within the space of possible phenotypes, developmental constraints can ``bias\ldots the production of variant phenotypes or {[}place{]} a limitation on phenotypic variability'' (\protect\hyperlink{ref-maynard_smith_developmental_1985}{Maynard Smith et al. 1985}). For example, Fibonacci phyllotaxis may arise from a packing constraint of primordia on the developing apex (\protect\hyperlink{ref-mitchison_phyllotaxis_1977}{Mitchison 1977}; \protect\hyperlink{ref-reinhardt_law_2022}{Reinhardt and Gola 2022}; but see \protect\hyperlink{ref-niklas_role_1988}{Niklas 1988} for an adaptive explanation). Natural selection constrains phenotypic variation by preventing maladaptive forms from evolving. Selection causes trait covariation because `missing' trait combinations are maladaptive. Understanding phenotypic constraint is challenging, but a useful starting point is determining whether phenotypic covariation can be explained by functional or developmental constraints (\protect\hyperlink{ref-mcghee_theoretical_1999}{McGhee 1999}, \protect\hyperlink{ref-mcghee_geometry_2007}{2007}; \protect\hyperlink{ref-olson_plant_2019}{Olson 2019}). If phenotypic covariation is inconsistent with functional and developmental constraints, this provides a strong impetus to test for selective constraint. Selective constraint does not mean that populations do not respond to selection, but rather that limits on phenotypic variation arise because selection never favors certain trait values.

In this study, we will address packing problems and developmental integration, specific forms of functional and developmental constraint relevant to our study system, stomatal anatomy. We introduce these concepts generally in this paragraph. Packing a number objects into a finite space is a common functional constraint on organisms. Regular geometries that appear in nature such as helices and hexagons (think DNA and honeycombs) are often optimal solutions to packing problems (\protect\hyperlink{ref-mackenzie_proving_1999}{Mackenzie 1999}; \protect\hyperlink{ref-maritan_optimal_2000}{Maritan et al. 2000}). Notice that functional constraint does not preclude selection, but the presence of a packing limit changes the range of possible phenotypes. Developmental integration is a form of developmental constraint on multivariate phenotypic evolution and we use these terms interchangeably in this study. Developmentally integrated traits have a ``disposition for covariation'' (\protect\hyperlink{ref-armbruster_integrated_2014}{Armbruster et al. 2014}), meaning that evolutionary divergence between lineages in one trait will be tightly associated with divergence in another trait. Allometry is a classic, albeit contested, example of developmental integration that may constrain phenotypic evolution (reviewed in \protect\hyperlink{ref-pelabon_evolution_2014}{Pélabon et al. 2014}). Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection (\protect\hyperlink{ref-lande_quantitative_1979}{Lande 1979}). However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection (\protect\hyperlink{ref-hansen_is_2003}{Hansen 2003}). For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators (Berg's rule, \protect\hyperlink{ref-berg_general_1959}{Berg 1959}, \protect\hyperlink{ref-berg_ecological_1960}{1960}; \protect\hyperlink{ref-conner_raissa_2014}{Conner and Lande 2014}; \protect\hyperlink{ref-armbruster_covariance_1999}{Armbruster et al. 1999}).

Biologists have studied phenotypic constraints for decades, but progress is challenging because many multivariate phenotypes are too complex or too poorly understood to quantitatively distinguish functional, developmental, and selective constraints over macroevolutionary timescales. Stomatal anatomy on the leaves of flowering plants provide an exceptional opportunity because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of years of evolutionary history. For this purpose, a heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selective constraints may differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selective constraints most likely explain at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological significance. The next sections provide background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important.

@@

Stomata are microscopic pores formed by a pair of guard cells that regulate gas exchange (CO\(_2\) gain and water vapor loss) on the leaves or other photosynthetic surfaces of most land plants. Stomata originated once in the history of land plants around 500 Ma, diversified rapidly in density and size, and have been maintained in most lineages except some bryophytes and aquatic plants (recently reviewed in \protect\hyperlink{ref-clark_origin_2022}{Clark et al. 2022}). Stomata respond physiologically by opening and closing in response to light, humidity, temperature, circadian rhythm, and plant water status (\protect\hyperlink{ref-hetherington_role_2003}{Hetherington and Woodward 2003}; \protect\hyperlink{ref-lawson_guard_2020}{Lawson and Matthews 2020}). The stomatal size, density, and distribution on a mature leaf do not change, so the maximum rate of gas exchange is fixed. However, the plant may respond plastically to environmental cues such as light and CO\(_2\) by altering stomatal anatomy in new leaves (\protect\hyperlink{ref-casson_influence_2008}{Casson and Gray 2008}). Physiological responses (aperture change) and plastic responses (new leaves with changed anatomy) may be alternative strategies for plants to acclimate to environmental change (\protect\hyperlink{ref-haworth_co-ordination_2013}{Haworth, Elliott-Kingston, and McElwain 2013}). Finally, stomatal anatomy can evolve due to inherited changes in stomatal development. Plastic and genetic changes in stomatal anatomy are both ecologically important, but most studies do not use a common garden design that would tease apart their relative contribution.

We focus on anatomical variation in the density, size, and patterning of stomata on a leaf because these factors set the maximum stomatal conductance to CO\(_2\) diffusing into a leaf and the amount of water that transpires from it (\protect\hyperlink{ref-sack_hydrology_2003}{Sack et al. 2003}; \protect\hyperlink{ref-franks_effect_2001}{Franks and Farquhar 2001}; \protect\hyperlink{ref-galmes_leaf_2013}{Galmés et al. 2013}; \protect\hyperlink{ref-harrison_influence_2020}{Harrison et al. 2020}). Plants typically operate below their anatomical maximum by dynamically regulating stomatal aperture. Even though operational stomatal conductance determines the realized photosynthetic rate and water-use efficiency, anatomical parameters are useful in that they set the range of stomatal function (\protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}) and are correlated with actual stomatal function under natural conditions (\protect\hyperlink{ref-murray_consistent_2020}{Murray et al. 2020}). All else being equal, larger, more densely packed, but evenly spaced stomata increase gas exchange (\protect\hyperlink{ref-franks_maximum_2009}{Franks and Beerling 2009}; \protect\hyperlink{ref-dow_physiological_2014}{Dow, Berry, and Bergmann 2014}; \protect\hyperlink{ref-lehmann_effects_2015}{Lehmann and Or 2015}). Smaller stomata may also be able to respond more rapidly than larger stomata, proving the ability of leaves to track short duration environmental change (\protect\hyperlink{ref-drake_smaller_2013}{Drake, Froend, and Franks 2013}). Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in some species (\protect\hyperlink{ref-metcalfe_anatomy_1950}{Metcalfe and Chalk 1950}; \protect\hyperlink{ref-parkhurst_adaptive_1978}{Parkhurst 1978}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}). Amphistomatous leaves have a second parallel pathway from the substomatal cavities through the leaf internal airspace to sites of carboxylation in the mesophyll (\protect\hyperlink{ref-parkhurst_adaptive_1978}{Parkhurst 1978}; \protect\hyperlink{ref-gutschick_photosynthesis_1984}{Gutschick 1984}). Thus amphistomatous leaves have lower resistance to diffusion through the airspace which increases the photosynthetic rate (\protect\hyperlink{ref-parkhurst_intercellular_1990}{Parkhurst and Mott 1990}). If total stomatal and other conductances to CO\(_2\) supply could be held constant, then an amphistomatous leaf will have a greater conductance than an otherwise identical hypostomatous leaf. The magnitude of the advantage depends on the resistance to diffusion through the internal airspace, which is variable among species.

The adaptive significance and ecological distribution of leaves with different stomatal anatomies is complex and there is much yet to learn. Seed plants posses a wider range of stomatal anatomies than ferns and fern allies, which are restricted to having large stomata, at low density, only on the lower surface (\protect\hyperlink{ref-de_boer_optimal_2016}{de Boer et al. 2016}). In general, trees and shrubs have greater stomatal density than herbs, but there is a of lot variation within growth forms depending on the ecological niche (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-kelly_plant_1995}{Kelly and Beerling 1995}). A commonly observed trend is that leaves from higher light environments tend to have greater stomatal density (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}; \protect\hyperlink{ref-gibson_structure-function_1996}{Gibson 1996}; \protect\hyperlink{ref-smith_associations_1998}{W. K. Smith, Bell, and Shepherd 1998}; \protect\hyperlink{ref-jordan_using_2014}{Jordan, Carpenter, and Brodribb 2014}; \protect\hyperlink{ref-muir_making_2015}{Muir 2015}; \protect\hyperlink{ref-bucher_stomatal_2017}{Bucher et al. 2017}). This may explain why, perhaps counterintuitively, plants in dry environments tend to have more stomata. Drier habitats are more open, enabling plants with higher stomatal density to photosynthesize more when water is available, but close stomata during drought (\protect\hyperlink{ref-liu_variation_2018}{Liu et al. 2018}). Over recent human history, stomatal density has tended to decline within species as atmospheric CO\(_2\) concentrations have risen (\protect\hyperlink{ref-woodward_stomatal_1987}{Woodward 1987}; \protect\hyperlink{ref-royer_stomatal_2001}{Royer 2001}). It is unclear whether most of this change is plastic or genetic, but the overall direction is consistent with the hypothesis that plants decrease gas exchange as CO\(_2\) availability increases.
